Site-Heterogeneous Character Change Models for Morphology

### 2018-10-29 Papers/DraftOutline_SSCCM.tex
adding some math

\documentclass[]{article}

@@

>>>>

\documentclass[]{article}
\usepackage{amsmath}

@@

\subsection{Modeling site-heterogeneous state frequencies}

\subsubsection{Discretized beta model}
We model site-specific state frequencies by assuming the state frequency $\pi_i$ at site $i$ is drawn from a Beta prior distribution with hyperparameters $\alpha,\beta$.
Then, the likelihood of the data $D$ is computed by marginalizing over the value for $\pi_i$ at each site.

\begin{equation}\label{continuous-likelihood}
f(D\mid \alpha,\beta) = \prod_{i=1}^n\int_0^1 f(D_i \mid \pi_i)f(\pi_i \mid \alpha,\beta)d\pi_i
\end{equation}

To simplify the computation, we approximate this integrated likelihood by assuming $\pi_i$ is drawn from a mixture distribution, where the mixture categories are defined deterministically by discretizing a Beta distribution into $k$ bins whose boundaries are defined using $k-1$ quantiles.
The the value of mixture category $j$ is indicated by $\beta_j$ and is computed as the interquantile median of the $j$-th bin.
In other words, $\beta_j = I^{-1}_{(2j-1)/2k}(\alpha,\beta)$, where $I_x(\alpha,\beta)$ is the regularized incomplete Beta function.
The mixing proportion is $1/k$ for each category.
Then the likelihood is computed by summing over the $k$ discrete mixture categories for $\pi_i$ at each site.

\begin{equation}\label{discretized-likelihood}
f(D\mid \alpha,\beta) = \prod_{i=1}^n\frac{1}{k}\sum_{j=1}^k f(D_i \mid \pi_i = \beta_j)
\end{equation}

Note that in the limit as the number of mixture categories approaches infinity, equation \ref{discretized-likelihood} is identical to equation \ref{continuous-likelihood}.

\subsubsection{Site-heterogeneous discrete multistate model}

When analyzing multistate character data with $S$ character states, we assume the state frequencies $\pi_i$ at each site $i$ are drawn from a Dirichlet distribution with concentration parameters $\alpha_1,\ldots,\alpha_S$.
Again, we use a mixture distribution to simplify the computation, but instead of defining the mixture categories using a discretized Dirichlet distribution, the value $\delta_j$ of mixture category $j$ is assumed to be drawn from the same underlying Dirichlet distribution with parameters $\alpha_1,\ldots,\alpha_S$.
The mixing proportions are $\theta_j$, which are themselves drawn from a uniform Dirichlet prior distribution.

$$f(D\mid \delta_1,\ldots,\delta_k,\theta_1,\ldots,\theta_k) = \prod_{i=1}^n \sum_{j=1}^k  f(D_i \mid \pi_i = \delta_j)\theta_j$$
